---
title: "STAT 340 Final exam - Question 3"
author: "Sean Wells"
date: "May 13, 2022"
output: html_document
---

```{r include=F}
knitr::opts_chunk$set(echo=T,warning=F,message=F,fig.align="center",error=T)
library(tidyverse)
library(lme4)
library(glmnet)
```



***REMEMBER:***


 - Unless manual/by hand calculation specifically requested, you can **use any R functions covered in class**. Functions not covered in class MAY NOT be graded (depending on context).
 - **All plots MUST have reasonably good titles/labels** (quality of titles/labels subject to determination by graders). Failure to do so MAY result in penalties.
 - Also note that `error=TRUE` has been turned on in the setup chunk. This means that **the file will knit EVEN if there are errors**! This will ensure you always have an HTML file to submit, but remember to ***CHECK YOUR FILE FOR ERRORS BEFORE YOU SUBMIT!!*** You can turn this off if you like, but it seems to help most people.



## Question 3: Great Crested Newt


A survey of great crested newts was conducted in the UK in an effort to better understand their ecology and conserve their habitats.^[https://www.arguk.org/info-advice/advice-notes/9-great-crested-newt-habitat-suitability-index-arg-advice-note-5/file] Many ponds were carefully examined and checked for presence of newt life (0 indicating no newts were present; 1 indicating at least some newts were present). Several predictors were also recorded for each pond, listed below:

 - `log10area`: pond area (m²)
 - `dry`: how often the pond dries (1-4: never, rarely, sometimes, annually)
 - `water`: water quality (1-4: bad, poor, moderate, good)
 - `shade`: % of shoreline in shade
 - `bird`: evidence of bird impact (1-3: none, minor, major)
 - `fish`: evidence of fish impact (1-4: none, possible, minor, major)
 - `ponds`: # of ponds within 1km
 - `land`: land habitat quality (1-4: none, poor, moderate, good)
 - `macro`: % pond area covered by plants

Here's the dataset. We're going to explore different ways of predicting the presence of newts in ponds.

```{r}
newt = read_csv("https://www.toptal.com/developers/hastebin/raw/ihemihizal")
head(newt)
```



### Part I <small>(pts: 1, 2, 2, 2)</small>


a. Create an 80/20 train/test split (i.e. randomly sample 80% of the rows to be used as a training data set, and set aside the remaining 20% as a testing data set).

```{r}
train = sample_n(newt, 160)
model = lm(formula = presence ~ ., family="binomial", data= train)
```



b. Start by fitting a full logistic regression model (i.e. with all terms). **Show a summary table** of the resulting fit

```{r}
summary(model)
```

c. Which predictors are significant? Write interpretations for the 3 most significant predictors. Be sure to indicate the following:
   - **the direction** of the relationship (i.e. is it linked to increasing or decreasing chance of newt presence?)
   - **the magnitude** of the relationship (i.e. how much are the chances increasing? Please use **precise** quantitative, statistical language).
   
It seems that the three most significant predictors are shade, ponds, and macro. 
Shade: For each percent increase in shade coverage of the shoreline, the chance of newt life being present in the pond decreases by 29%. 
Ponds: For each increase in the number of ponds within 5 km, the chance of newt life being present in the pond increases by 4.6%.
Macro: For each percent increase in the percentage of the pond covered in plants, the chance of newt life being present in the pond increases by 35%.

d. Use your fitted model to generate predictions on the test data set and compare your predictions to the true test responses to **estimate the out-of-sample mean squared error** (MSE) of this model. **Report the MSE**. (*Hint:* remember to set the `type` argument appropriately to get actual probabilities).

```{r}
test = anti_join(newt, train)
test$test = predict(model, test)
test
```

### Part II <small>(pts: 3, 1, 2)</small>


Let's try a different approach where we try to **simultaneously perform regularization and variable selection**.

a. Choose a regularization method that is also good at selecting out a smaller subset of variables (*Hint:* which method is good at setting coefficients to exactly 0?). Using the `glmnet` package, fit a model of this type, making sure to set the correct value for `alpha`. Also add the argument `family=binomial` to ensure you get a logistic-type fit.

```{r}
nrows <- nrow(newt); # Number of observations in the data
errors <- data.frame( 'Row'=rep(1:nrows),
                      'Error'=rep(NA, nrows));

for ( i in 1:nrow(newt) ) {
  train_data <- newt[-c(i),]; # Leave out the i-th observation
  leftout <- newt[c(i),]; # the row containing the left-out sample.
  
  # Fit the linear model, then evaluate.
  x = model.matrix(presence~ .,train_data)[,-1]
  y = train_data$presence
  lm.lasso = glmnet(x,y,alpha=1)
  
  newt.pred <- predict(lm.lasso, s=0.25, newx=model.matrix(presence~., leftout)[,-1]);
  idx <- (errors$Row==i); # Pick out row of the errors df.
  # record squared error btwn predict and truth
  errors[idx,]$Error <- (newt.pred - leftout$presence)^2 ; }
mean(errors$Error)
```

b. Show the coefficients of your fit. Which predictors were removed from the model by your method? 

```{r}
coef(lm.lasso)
```

Dry and bird were removed most commonly. 

c. Again, use your fitted model to generate predictions on the test dataset and compare your predictions to the true test responses to **estimate the out-of-sample mean squared error** (MSE) of this model. **Report the MSE** (*hint:* remember to set `type` appropriately!), and **compare this value to the MSE from the first model**. Is this MSE higher or lower? By what percent approximately?

The true out of sample MSE is approximately 0.23. 

```