
---
title: "STAT340 HW2: Monte Carlo & Testing"
author: Sean Wells
date: Mar 4, 2022
output: html_document
---
<style>
table{width:50%!important;margin-left:auto!important;margin-right:auto!important;}
/*ol[style*="decimal"]>li{margin-top:40px!important;}*/
</style>

```{r setup, include=FALSE}
# check packages installed
if(!require(pacman)) install.packages("pacman")
pacman::p_load(ggplot2)

knitr::opts_chunk$set(tidy=FALSE,strip.white=FALSE,fig.align="center",comment=" #")
options(width=100)
```

<br/>

## Instructions

Complete the exercises, update the "author" and "date" fields in the header, knit it, and submit **both the HTML and RMD** files to Canvas. Due date: **Mar 4, 2022 at 11:59pm**.

---

## Exercise 1 (20 points): Generalized [birthday problem](https://en.wikipedia.org/wiki/Birthday_problem)

The birthday problem asks for the probability that in a group of $n$ people, at least 2 people will share the same birthday. This is easy to solve, and the solution is easily found online.

We can generalize this to a more difficult problem and solve it using a Monte Carlo approach: in $n$ people, what is the probability that at least $k$ people have the same birthday?

Write a function `birthday(n,k,i)` that returns a probability estimate given 3 arguments:

 - $n$ is the number of people in your sample
    - for example, if `n=50` is used, we are asking "in 50 people, what is the probability that..."
 - $k$ is minimum number of people that must share a birthday
    - for example if `k=4` is used, we asking "...what is the probability that at least 4 people share the same birthday?
 - $i$ is the number of iterations to run (default 1000)
    - for example, if `i=1000` is used, your function should run 1000 simulations

**Notes**:

 - You may assume there are 365 possible dates (no leap years)
 - You may assume birthdays are uniformly distributed across the calendar
    - this is actually not true, see [this](https://www.panix.com/~murphy/bday.html), or [this](https://fivethirtyeight.com/features/lots-of-parents-dont-want-their-kids-to-be-born-on-leap-day/)
 - You may assume the people are sampled [i.i.d](https://en.wikipedia.org/wiki/Independent_and_identically_distributed_random_variables)

**Hints**:

 1. There's no need to use actual dates in the simulation process. Numbers can represent dates and are easier to generate and manipulate in `R`. In particular, we recommend using the `sample()` function with the `x`, `size`, and `replace` arguments set appropriately. See the help page `?sample` for details.
 2. Given a vector of numbers, you can easily find duplicates by using the `table()` function. This will produce a named vector showing how many of each value there are. For example, running `table(c(1,3,5,5,7,9,9,9))` will show you there is one 1, one 3, two 5s, one 7, and three 9s.
 3. In your function, you will need to use a `for` loop to repeat the simulation `i` times. You will also need a variable outside your `for` loop to keep track of how many simulations satisfy that \# of birthdays $\geq k$.
 4. If your function is running correctly, then `birthday(n=23, k=2)`, `birthday(n=87, k=3)` and `birthday(n=188, k=4)` should all be approximately $50\%$.
 5. If your function is very slow, consider using the [`Table` function](https://rdrr.io/cran/Rfast/man/Table.html) from the Rfast package, which is 4-5 times faster than the normal `table()` function.

```{r include=FALSE}
library(tidyverse)
# complete the function
# note i=1000 sets the default value of i to be 1000
birthday = function(n,k,i=1000){
  shared = 0
  for (i in 1:i) {
    birthdays = sample(x = 1:365, size = n, replace = TRUE)
    shared = shared + sum(max(table(birthdays)) >= k)
  }
  shared / i
}
```

This class currently has 162 enrolled students. What is the approximate probability that at least $4$ students have the same birthdays?

> **0.335**

---

## Exercise 2 (15 points): Simulate RV

$X$ is a random variable defined between $-\frac\pi2$ and $\frac\pi2$ by the probability density function $f(x)=\frac12\cos(x)$. Note this means the cumulative distribution function is $$F(x)=\int_{-\frac\pi2}^xf(x)dx=\frac12(\sin(x)+1)$$ Write a function `rx(n)` to sample from this random variable, where `n` is the size of the sample to be drawn. Then, use your function to draw a sample of 1000 and plot a histogram of the output.

**Hint:** Remember the inverse function of $\sin(x)$ is $\arcsin(x)$, which in `R` is found by using the built in function `asin()`

```{r,fig.width=4,fig.height=3}
# defining pdf of X
pdf_x = Vectorize(function(x){
  if(x>-pi/2 & x<pi/2){cos(x)/2} else 0
})

# showing pdf on plot
ggplot() + geom_function(fun=pdf_x,n=10001) + theme_minimal() + 
  xlim(c(-2.5,2.5)) + ylim(-.5,1) + labs(x='x',y='f(x)')
```

```{r}
# complete the function
rx = function(n){
  x.unif = runif(n,0,1)
  x = asin((2 *x.unif) - 1)
}

# uncomment the following line of code and check it looks correct
hist(rx(1000))
```

---

## Exercise 3 (15 points): Testing coin flips

In the six sequences below, **only one** of them is actually randomly generated from a fair coin. Use a combination of everything you know (common sense, monte carlo, hypothesis testing, etc.) to identify which is actually random and explain your reasoning.

(***For full points***, conduct a formal test and report a $p$-value for each sequence. You may use a combination of multiple tests to arrive at your answer. If you cannot compute a $p$-value for each sequence, you can still earn a significant amount of partial credit by carefully explaining your reasoning and response as best as you can.)

```{r}
flips1 = "HTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHT"

flips2 = "HHHTHTTTHHTHHTHHHTTTTHTHTHHTTHTHHHTHHTHTTTHTHHHTHTTTHTHTHHTHTHTTHTHHTHTHTTTHTHHHTHTHTTHTHTHHTHTHTHHHTHTTTHTHHTHTHTHHTTTHTHHTHHTTTTHTHTHHHTHTTHTHHTHTHTTHTHHTHTHHHTHHHTHTTTHTTHTTTHTHHHTHTHTTHTHHTHHTHTTT"

flips3 = "HHTHTHTTTHTHHHTHHTTTHTHHTHTTTHTHTHHTHTHTTHTHHHHHHTTTHTHTHHTHTTTHTHHTHTHTTTHTHHHTTHTTTHTHTHHHHTHTTHHTTTTTHTHHHTHTHTTTTTHHHTHHTHHTHHHTTTTHTHTHHHTHHTTTTTHTHHHTHTHTHTTTHTHHHTHTHTHTTHTHHTHTHTHTTTTHTHHHTHTH"

flips4 = "HTHHHHHHHTHTTHHTTHHHTHTHTTTHHTHHHTHHTTHTTTTTTTTTHTHHTTTTTHTHTHTHHTTHTTHTTTTTHHHTHTTTHTHTHHHTHTTTTHTHTHHTTHTHTTHHTHTHHHHTHTTHHTTHTTHTTHTHHHHHHTTTTTTHHHTTHTHHHHTTTHTTHHHTTHTHHTTTHHTHHTTTHTHHTHHHTHHTTHHH"

flips5 = "HHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTT"

flips6 = "TTHTTTHTTTTTTTHTHTHTHTTHTTHTHHTHHTTTHHTHTTTHTHHTHHHTHTTHHTHHTTHTHTTTTHTHTTTHHTTTTTTTTHTHHTTHTTTTTTHTHTHTHTTTHTTHHTTHTTTHHTTTHTTHTTTTHTTTTHHTTTHTHTHHHTTTTTTHTHHTTTTTTTTTTTTHHHTTTHHHTTTHTTTHTHTTHTTTTTHT"

# you can use the function below to split the above sequences in vectors of flips
split = function(str) strsplit(str, split="")[[1]]
split(flips1)
```

H0 : Each coin flip is independent Bernoulli(p=1/2)

```{r} 
split1 = split(flips1)
split2 = split(flips2)
split3 = split(flips3)
split4 = split(flips4)
split5 = split(flips5)
split6 = split(flips6)

#from 04_testing.rmd

longestRun = function(flips){
  MaxRunSoFar = 1
  currentRun = 1
  for(i in 2:length(flips)){ 
    if(flips[i]==flips[i-1]){
      currentRun = currentRun + 1
      if(currentRun>MaxRunSoFar) MaxRunSoFar = currentRun
    }
    if(flips[i]!=flips[i-1]){
      currentRun=1
  }
  }
  return(MaxRunSoFar)
}

simulate_S = function(){
  S = sample(c("H","T"), 200,replace=T) %>% longestRun()
  return(S)
}

r = 1000
monte_carlo = data.frame(replicate = 1:r, 
                         S = rep(NA,r), 
                         S_in_A = rep(NA, r)) 

for(i in 1:r){
  monte_carlo$S[i] = simulate_S()
}
hist(monte_carlo$S, breaks = 20)
lines(c(1,1), c(0,100000), col = "red", lwd = 3)
lines(c(4,4), c(0,100000), col = "red", lwd = 3)
lines(c(6,6), c(0,100000), col = "red", lwd = 3)
lines(c(9,9), c(0,100000), col = "red", lwd = 3)
lines(c(10,10), c(0,100000), col = "red", lwd = 3)
lines(c(12,12), c(0,100000), col = "red", lwd = 3)
```
```{r}
check_if_S_in_A = function(S){
  return(S  <= 1)
}

for(i in 1:r){
  monte_carlo$S_in_A[i] = check_if_S_in_A(monte_carlo$S[i])
}

monte_carlo = as_tibble(monte_carlo)
p_value1 = monte_carlo %>% summarise(mean(S_in_A))
p_value1
```

```{r}
check_if_S_in_A = function(S){
  return(S  <= 4)
}

for(i in 1:r){
  monte_carlo$S_in_A[i] = check_if_S_in_A(monte_carlo$S[i])
}

monte_carlo = as_tibble(monte_carlo)
p_value2 = monte_carlo %>% summarise(mean(S_in_A))
p_value2
```

```{r}
check_if_S_in_A = function(S){
  return(S  <= 6)
}

for(i in 1:r){
  monte_carlo$S_in_A[i] = check_if_S_in_A(monte_carlo$S[i])
}

monte_carlo = as_tibble(monte_carlo)
p_value3 = monte_carlo %>% summarise(mean(S_in_A))
p_value3
```

```{r}
check_if_S_in_A = function(S){
  return(S  <= 9)
}

for(i in 1:r){
  monte_carlo$S_in_A[i] = check_if_S_in_A(monte_carlo$S[i])
}

monte_carlo = as_tibble(monte_carlo)
p_value4 = monte_carlo %>% summarise(mean(S_in_A))
p_value4
```

```{r}
check_if_S_in_A = function(S){
  return(S  <= 10)
}

for(i in 1:r){
  monte_carlo$S_in_A[i] = check_if_S_in_A(monte_carlo$S[i])
}

monte_carlo = as_tibble(monte_carlo)
p_value5 = monte_carlo %>% summarise(mean(S_in_A))
p_value5
```

```{r}
check_if_S_in_A = function(S){
  return(S  <= 12)
}

for(i in 1:r){
  monte_carlo$S_in_A[i] = check_if_S_in_A(monte_carlo$S[i])
}

monte_carlo = as_tibble(monte_carlo)
p_value6 = monte_carlo %>% summarise(mean(S_in_A))
p_value6
```

The Monte Carlo probabilities for flips 1 and 2 are statistically significant; contain longest runs that are improbable under the null hypothesis. Thus, for these two trials we reject the null hypothesis that they were randomly generated from a fair coin. We fail to make any conclusions for the other four trials based upon the results of this test however. 

To obtain further conclusions, we can examine the odds ratios of the sequences to suggest dependence vs independence for each trial. A R ratio close to 1 suggests independence between each trial, supporting the null hypothesis that each trial was conducted with use of a fair coin. 

```{r}
tableOfPairs = function(vec){
  return(table(runner::runner(vec,k=2,f=paste,collapse="")[-1]))
}

ratioFromTable = function(tb){
  return(setNames((tb["HH"]/tb["HT"])/(tb["TH"]/tb["TT"]),"R"))
}
```

```{r}
ratioFromTable(tableOfPairs(split3))
```

```{r}
ratioFromTable(tableOfPairs(split4))
```

```{r}
ratioFromTable(tableOfPairs(split5))
```

```{r}
ratioFromTable(tableOfPairs(split6))
```

The results of the odds ratios suggest that in flips 3 and 6, subsequent heads were less likely to happen after a heads was landed, whereas in flip 5 a heads was significantly more likely to happen after a heads occurred. Flip 4 has an odds ratio closest to 1 which suggests that this sequences was the most likely to have been conducted with the use of a fair coin. 

As a side note, sequence 1 and 5 could be ruled out from the start seeing as the first sequence is simply alternating heads and tails whereas sequence 5 is strictly alternating streaks of ten heads/tails; both of which are highly unlikely to happen naturally (via use of a fair coin).

Response goes here: The results of both Monte Carlo probability testing and the use of odds ratios suggest that sequence #4 was the only sequence conducted with the use of a fair coin and was truly randomly generated. 
